Inferring the gravitational potential of the Milky Way with a few precisely measured stars

### 2014-02-13 text/paper2/ms.tex
rambling

Cosmological simulations of galaxy formation in the $\Lambda$CDM paradigm predict dark matter halos that (1) are permeated with substructure on many scales, (2) are triaxial in shape, and (3) have density profiles and shapes that vary with radius \citep{dubinski91, jing02, kuhlen07, veraciro11}. Dark-matter-only simulations produce strongly triaxial halos \citep{} with large density variations \citep{zemp09}. Including baryons tends to soften the triaxiality in the inner galaxy through energetic feedback \citep[e.g.,][]{pontzen12} but does not erase the clumpy, triaxial nature of the outer halo. This leads to radially-dependent axis ratios and orientation, density profiles, and smoothness, thus predicting complicated, non-intregrable potential forms for Milky Way-like galaxies.

The bulk of the baryonic matter in galaxies spans roughly 5-10\% of the extent of the host dark matter halo. Hence, the major components of a galaxy are sensitive to the inner portion of its host halo's mass distribution. For example, the rotation curves of disk galaxies are particularly sensitive probes of the inner mass since matter in disks can be assumed to move on near circular orbits. This extra knowledge about the orbital distribution of the baryonic matter within a disk allows a more precise tracing of the potential than is feasible for objects such as elliptical galaxies where orbits are assumed to be random. [Potential is most likely smoothed out (energetic feedback), can approximate as closely integrable allowing beautiful dynamical methods ...]

Conversely, measuring the dark matter distribution at large radii is difficult as the density of visible tracers becomes small. Around external galaxies, the extended mass distribution has been explored using a variety of approaches \citep[see][for a a complete and detailed review]{courteau12}. For example, the kinematics of tracer populations such as globular clusters or planetary nebulae can be used to derive mass estimates under the assumptions that these satellite systems are relaxed and well-mixed \citep[early investigations include][]{mendez01,cote03}. These estimates are necessarily limited by the number of objects bright enough for spectroscopic measurements and thus the largest samples to date have roughly several hundred members probing out to several 10's of kpc. Simple, parameterized models of both the mass and orbit distribution have been simultaneously constrained using such data \citep[e.g.][]{napolitano11,deason12a}. 
%\citep{romanowsky03,douglas07,romanowsky09,napolitano11,lee11}
Alternatively, the statistical properties of gravitationally lensed background sources around a galaxy can be used to constrain the \emph{projected} shape, orientation, and radial profile of mass \citep[as done by the Lens Structure and Dynamics Survey described in][]{koopmans02}. Of course, lensing reconstructions can only be performed for galaxies which closely intersect our line of sight to background sources, but the advent of large photometric catalogues has allowed automatic searches for such chance alignments and significant increases in the number of objects studied in this way \citep[e.g. the Sloan Lens ACS Survey, see][]{bolton06}.

Within the Milky Way we have a unique vantage point that allows us to measure the three-dimensional structure of our own dark matter halo. One approach to ... [explain ``random'' methods: kinematic tracers]. Another approach is to exploit the non-random nature of the stellar populations in the halo. A consequence of hierarchical structure formation [...] and such structures have been observed [e.g., ...cite belokurov?]. Tidal streams are dynamically cold systems --- debris typically have small distributions of energy and angular momentum --- and thus require orders of magnitude fewer tracers than a random sample to get constraints of comparable accuracy. For example, in the simplest case we might {\it assume} that debris stars are actually still on the same orbit as their progenitor system \citep[a \emph{wrong} assumption, see e.g.][]{eyre}. Then we can imagine measuring the full-space velocities ${\bf v}$ at different points ${\bf x}$ in the structure (e.g. along a stream), which would give us a direct measure of differences in a potential, $\Phi$. In practice, simple methods such as orbit fitting lead to systematic biases in inferred properties of the underlying potential because streams do not strictly delineate orbits [cite eyre binney work].

>>>>

Cosmological simulations of galaxy formation in the $\Lambda$CDM paradigm predict dark matter halos that (1) are permeated with substructure on many scales, (2) are triaxial in shape, and (3) have density profiles and shapes that vary with radius \citep{dubinski91, jing02, kuhlen07, veraciro11}. Dark-matter-only simulations produce strongly triaxial halos \citep{} with large density variations \citep{zemp09}. Including baryons tends to soften the triaxiality in the inner galaxy \citep[e.g.,][]{dubinski94, pontzen12} but does not erase the clumpy, triaxial nature of the outer halo. This leads to radially-dependent axis ratios and orientation, density profiles, and smoothness, thus predicting complicated, non-intregrable potential forms for Milky Way-like galaxies.

The bulk of the baryonic matter in galaxies spans roughly 5-10\% of the extent of the host dark matter halo. Hence, the major observable components of a galaxy are sensitive to the inner portion of its host halo's mass distribution. For example, the rotation curves of disk galaxies are excellent probes of the inner mass, especially since matter in disks can be assumed to move on near circular orbits. In the inner regions of the Galaxy, the potential is likely modified due to past dissipative infall \citep{dubinski94} or cooling \citep{bryan13} of baryons, which act to reduce triaxiality and smooth out density fluctuations. The inner potential is likely well-described by a closely-integrable potential, thus allowing [...]

%This extra knowledge about the orbital distribution of the baryonic matter within a disk allows a more precise tracing of the potential than is feasible for objects such as elliptical galaxies where orbits are assumed to be random.

Conversely, measuring the dark matter distribution at large radii is complicated by the low density of visible tracers and [...something about assumptions about integrability and smoothness]. Around external galaxies, the extended mass distribution has been probed using a variety of approaches \citep[see][for a a complete and detailed review]{courteau12}. For example, the kinematics of tracer populations such as globular clusters or planetary nebulae can be used to derive mass estimates under the assumptions that these satellite systems are relaxed and well-mixed \citep[early investigations include][]{mendez01,cote03}. These estimates are necessarily limited by the number of objects bright enough for spectroscopic measurements and thus the largest samples to date have roughly several hundred members probing out to several 10's of kpc. Simple, parameterized models of both the mass and orbit distribution have been simultaneously constrained using such data \citep[e.g.][]{napolitano11,deason12a}. 
%\citep{romanowsky03,douglas07,romanowsky09,napolitano11,lee11}
Alternatively, the statistical properties of gravitationally lensed background sources around a galaxy can be used to constrain the \emph{projected} shape, orientation, and radial profile of mass \citep[as done by the Lens Structure and Dynamics Survey described in][]{koopmans02}. Of course, lensing reconstructions can only be performed for galaxies which closely intersect our line of sight to background sources, but the advent of large photometric catalogues has allowed automatic searches for such chance alignments and significant increases in the number of objects studied in this way \citep[e.g. the Sloan Lens ACS Survey, see][]{bolton06}.

Within the Milky Way our unique vantage point allows us a three-dimensional view of the structure in own dark matter halo. Deason et al. (201?) used BHB stars selected to be in the halo from the Sloan Digital Sky Survey \cite[SDSS;][]{} as random tracers [...reconstruct velocity trends?]. Another approach is to exploit the non-random nature of the stellar populations in the halo. A consequence of hierarchical structure formation [...] and such structures have been observed [e.g., ...cite belokurov?]. Tidal streams are dynamically cold systems --- debris typically have small distributions of energy and angular momentum --- and thus require orders of magnitude fewer tracers than a random sample to get constraints of comparable accuracy. For example, in the simplest case we might {\it assume} that debris stars are actually still on the same orbit as their progenitor system \citep[a \emph{wrong} assumption, see e.g.][]{eyre}. Then we can imagine measuring the full-space velocities ${\bf v}$ at different points ${\bf x}$ in the structure (e.g. along a stream), which would give us a direct measure of differences in a potential, $\Phi$. In practice, simple methods such as orbit fitting lead to systematic biases in inferred properties of the underlying potential because streams do not strictly delineate orbits [cite eyre binney work].
